Towards atomistic resolution structure of phosphatidylcholine glycerol backbone and choline headgroup at different ambient conditions

### 2015-01-20 HGmodel_draft3.tex
new version

  \begin{figure}[]
  \centering
%  \includegraphics[width=8.6cm]{POPCstructure.eps}

@@

\sout {Berger-POPC-07~\cite{ollila07a}}\todoi{Not included in the Fig. 2, to be removed}          &   POPC & 128 & 7290  & 323  & 56 & 56  & ? & SI \\

@@

CHARMM36\cite{klauda10}\todoi{Hubert Santuz, these files are already in Gitbub, however to get doi citation it should be put to Zenodo}        & POPC   & 128 &  5120    & 303 & 150 & 100  & ? & SI \\
CHARMM36\cite{klauda10}\todoi{Samuli, put to Zenodo}        & DPPC   & 72  &  2189 & 323 & 30 & 25  & ? & SI \\
CHARMM36\cite{klauda10}\todoi{Not included in the Fig. 2. Should we remove or include?}        & DOPC   & 128 &  5120    & 303 & 150 & 100  & ? & SI \\

@@

Poger\cite{poger10}             & DPPC  & 128 & 5841 & 323 & 2$\times$100 & 2$\times$50 & \cite{pogerFILES} & SI \\

@@

\subsection{Full hydration: Experimental order parameters for the glycerol backbone and headgroup}\label{experiments}
The specific deuteration of $\alpha$-, $\beta$- and g$_3$- segments of dipalmitoylphosphatidylcholine (DPPC) has been successful, 
allowing the order parameter measurements for these segments by $^2$H--NMR~\cite{gally75,brown77,brown78,akutsu81}.
In addition, the order parameters for all glycerol backbone and choline headgroup segments in egg yolk lecithin~\cite{hong95a},
1,2-dimyristoyl-sn-glycero-3-phosphocholine (DMPC)~\cite{hong95b,gross97,dvinskikh05a}, 
1,2-dioleoyl-sn-glycero-3-phosphocholine (DOPC)~\cite{warschawski05} and POPC~\cite{warschawski05,ferreira13}
have been measured with several different implementations of $^{13}$C NMR experiments.
The experimental absolute values of glycerol backbone and choline order parameters from various publications are shown in Fig.~\ref{HGorderparameters}.
\begin{figure}[]
  \centering
%  \includegraphics[width=8.6cm]{comparison3.pdf}
%\todo{DONE}\\
%The data in this figure should be updated to be consistent with the systems listed in Table~\ref{systems}. At least, Tjornhammar et al and CHARMM-UA values are missing. Some results in the figure are outdated due to longer simulations reported etc.} \\
\todo{Markus Miettinen suggested that we should consider making one more figure where only experimental data would be shown and that would be discussed in Section~\ref{experiments}} \\
\todo{There has been some discussion in the blog that maybe we should include the signs and stereospecific labeling here as well, and that maybe this figure should be clarified. My view is that for the main conclusion, i.e. CHARMM36, GAFFlipid and MacRog are the best here, this format is good enough. However, if someone wants to improve this and/or collect all the data with the signs and/or stereospecific labeling it would be even better. There is also very interesting discussion about interactive figure.}
  \caption{\label{HGorderparameters}
  Order parameteres from simulations listed in Table~\ref{systems} and experiments for glycerol and choline groups.
  The experimental values were taken from the following publications: DMPC 303~K from \cite{gross97}, DMPC 314~K from \cite{dvinskikh05a}, DPPC 310K from \cite{gally75}, 
  DPPC 323~K from \cite{akutsu81}, and POPC 298~K \cite{ferreira13}.
  The vertical bars shown for some of the computational values are not error bars, but demonstrate that for 
  these systems we had two data sets; the ends of the bars mark the values from the two sets, and the dot marks their measurement-time-weighted average. 
  The H{\"o}gberg et al. force field in Table~\ref{systems} is abbreviated in this figure as HoNiLy, Tj{\"o}rnhammar et al. as TjoEdh, CHARMM36-UA as C36UA,
  Rabinovic et al. as RabLyu and Ulmschneider and Ulmschneider as UlmUlm.
} 
\end{figure}

@@

By combining the experimental information of the sign~\cite{hong95a,hong95b,gross97} and the stereospecific
measurements~\cite{gally81} with the accurate absolute value measurements from various techniques~\cite{gally75,akutsu81,gross97,dvinskikh05a,ferreira13},
the most detailed experimentally available order parameter information for the glycerol backbone and choline segments of POPC is obtained.
These data are shown in Fig.~\ref{HGorderparameters2}.
\begin{figure*}[]
  \centering
%  \includegraphics[width=8.6cm]{HGorderparameters5.eps} \\
  \todo{Markus Miettinen suggested that we should consider ``spreading'' the values similarly to the figure~\ref{HGorderparameters}} \\
  \caption{\label{HGorderparameters2}
  Order parameteres from simulations with Berger-POPC-07, CHARMM36, GAFFlipid and MacRog force fields together with experimental values for POPC glycerol and choline groups.
  The magnitudes for experimental order parameters are taken from Ferreira et al.~\cite{ferreira13}, the signs are based on the measurements by Hong et al.~\cite{hong95a,hong95b} 
  and Gross et al.~\cite{gross97}, and the R/S labeling is based in the measurements by Gally et al.~\cite{gally81}.
} 
\end{figure*}

@@

The dihedral angle distributions for the  g$_3$-g$_2$-g$_1$-O(\textit{sn}-1) dihedral calculated from different models are
shown in Fig.~\ref{dihDISTS}. The distribution is qualitatively different for the Berger-POPC-07 model, showing a maximum in 
the gauche$^+$-conformation (60$^o$) compared to all the other models showing a maximum in the anti-conformation (180$^o$).
The distributions in all the other models have the same general features, the main difference being that the
fraction of configurations in the gauche$^-$-conformation (-60$^o$) is zero for the MacRog, detectable for the CHARMM36 and
equally large to the gauche$^+$ fraction in GAFFlipid. From the results we conclude that most likely the wrongly sampled
dihedral angle for the g$_2$-g$_1$ bond explains the significant discrepancy to the experimental order parameters
for the g$_1$ segment in the Berger-POPC-07 model (Fig.~\ref{HGorderparameters2}). 
In conclusion, models preferring the anti conformation for this dihedral give more realistic order parameters and
this in agreement with previous crystal structure and $^1$H NMR studies~\cite{hauser80,hauser81,hauser81b,hauser88,pascher92,marsh06}.
\begin{figure}[]
  \centering
%  \includegraphics[width=8.6cm]{g1-g2_Cdihs2.eps}
  \caption{\label{dihDISTS}
    Dihedral angle distributions for g$_3$-g$_2$-g$_1$-O(\textit{sn}-1) dihedral from different models (POPC bilayer in full hydration).
      } 
\end{figure}

@@

%Interestingly, the probability of the gaughe conformations correlates with the order parameter difference between the $\beta$ and $\alpha$ segments:
%the larger the gaughe fraction the larger the order parameter difference. This suggestion together with the results in 
%Fig.~\ref{HGorderparameters2} would indicate that the correct gaughe-trans fraction for N-$\beta$-$\alpha$-O dihedral is 
%larger than in GAFFlipid but smaller than in CHARMM36. This happens to be quite close to the gaughe-trans fraction in
%the Berger model. However, there is significant forking and numerical values are off from the experiments in the Berger
%model suggesting that it has some other inaccuracies.
\begin{figure}[]
  \centering
%  \includegraphics[width=8.6cm]{a-bDIHS2.eps}
  \caption{\label{dihDISTS2}
    Dihedral angle distributions for N-$\beta$-$\alpha$-O dihedral from different models (POPC bilayer in full hydration).
  } 
\end{figure}

@@

\subsubsection{Phospholipid bilayer with low hydration level}
The experimental order parameters available in the literature~\cite{dvinskikh05b,ulrich94,bechinger91} 
for the glycerol backbone and choline as a function of hydration level are shown in Fig.~\ref{ordPhydr}. 
The independently reported values for choline segments are in good agreement with each other (despite 
slight differences in temperature and acyl chain composition),
showing increase for both segments with decreasing hydration level. It should be noted that only 
absolute values were measured in the original experiments~\cite{dvinskikh05b,ulrich94,bechinger91}, but
we have included the signs measured in separate studies~\cite{hong95a,hong95b,gross97}. 
Consequently, the $\beta$ order parameter with negative sign actually increases with dehydration 
since the absolute value decreases~\cite{dvinskikh05b,ulrich94,bechinger91}.
Slight decrease for the glycerol backbone g$_3$- and g$_2$- order parameters were observed with dehydration, 
while g$_1$ remained practically unchanged~\cite{dvinskikh05b}.
\begin{figure}[]
  \centering
%  \includegraphics[width=8.6cm]{OrderParameterDEHYDpresentation.eps}
\todo{Several people think that this plot should be clarified.}
  \caption{\label{ordPhydr}
    The effect of dehydration on glycerol and choline order parameters in experiments.
    The magnitudes of order parameters are measured for DMPC ($^{13}$C NMR) at 314~K~\cite{dvinskikh05b}, 
    for POPC ($^2$H NMR) at 296~K~\cite{bechinger91} and for DOPC ($^2$H NMR) at 303~K~\cite{ulrich94}. 
    The signs are based on the measurements by Hong et al.~\cite{hong95a,hong95b} 
    and Gross et al.~\cite{gross97}.
  }
\end{figure}

@@

The qualitative agreement with experiments in all simulation models for the $\alpha$ and $\beta$ order parameters  
as a function of hydration indicates that the structural response of the choline headgroup to dehydration is somewhat realistic
despite the unrealistic structures at full hydration. 
The most likely explanation is that the choline group
orients more parallel to the membrane plane with dehydration due to restricted interlamellar space. 
Indeed, the P--N (phoshate phosphorus to choline nitrogen) vector angle with membrane normal shows an increase for
all models as a function of dehydration in Fig.~\ref{PNangle}.
However, the qualitative agreement in the lipid response to dehydration does not guarantee the correct 
free energy landscape if the simulation model has an incorrect structure. The influence of this issue to 
dehydration energetics studied with simulations~\cite{eun09,schneck12} is left for future studies.
\begin{figure}[]
  \centering
%  \includegraphics[width=8.6cm]{PNangles.eps}

@@

The choline headgroup and glycerol backbone order parameters for POPC measured by $^{13}$C NMR~\cite{ferreira13} and DPPC choline order parameters 
measured by $^{2}$H NMR~\cite{brown78} are shown in Fig.~\ref{ordPchol} as a function of cholesterol content.
The agreement between different experimental results is again very good, showing only very modest changes in 
the choline order parameters as a function of cholesterol content. It should be noted, however, that very small
changes are measurable with high resolution $^{2}$H NMR experiments
and cholesterol causes a measurable increase in the $\beta$ order parameter and a forking in the $\alpha$ order
parameter~\cite{brown78}. These effects are, however, so small that they are barely visible in the scale used in Fig.~\ref{ordPchol}.
Further, the effects of cholesterol on the glycerol backbone order parameters for POPC from $^{13}$C NMR experiment~\cite{ferreira13} 
are in good agreement with the results for the phosphatidylethanolamine (PE) measured by $^{2}$H NMR~\cite{ghosh82}.
These results further support the idea that the glycerol backbone structural behaviour is independent of the
headgroup composition~\cite{gally81} and that the headgroup stucture is independent of the acyl chain region content unless
charges are present~\cite{scherer87}.
\begin{figure}[]
  \centering
%  \includegraphics[width=8.6cm]{OrderParameterCHOLexpVSsim.eps}
  \caption{\label{ordPchol}
    The effect of cholesterol content on the glycerol backbone and choline order parameters in experiments~\cite{brown78,ferreira13} and simulations
    with the Berger-POPC-07/H\"oltje-CHOL-13, CHARMM36 and MacRog force fields. The signs in the experimental values are based on the measurements by Hong et al.~\cite{hong95a,hong95b} 
    and Gross et al.~\cite{gross97}.  Most order parameters from Berger-POPC-07/H\"oltje-CHOL-13 model for g$_1$ are beoynd the y-axis scale.}
\end{figure}

@@

Since Poger lipids come from the GROMOS force field, it is important to note that GROMOS uses the RF scheme for computing electrostatics (this is the method used for the 
force field parameterization). Using setup i) based on RF, we were able to reproduce the results (i.e. area per lipid value of 0.63~nm$^2$) from the original work only with 
GROMACS versions 4.0.X and earlier (the original authors~\cite{poger10} used GROMACS version 3.3.3). When switching to versions 4.5.X and above, the area per lipid dropped to below 0.58~nm$^2$. 
The GROMACS developers were contacted and a redmine issue opened (\texttt{http://redmine.gromacs.org/issues/1400}). The difference comes from the new Trotter decomposition 
introduced in versions 4.5.X. A fix has been introduced in version 4.6.6 that allows a recovery of an area per lipid value of 0.615~nm$^2$. The results in terms of area per lipid using the different 
GROMACS versions are available at~\cite{fuchs14}.
% http://www.dsimb.inserm.fr/~fuchs/project\_Samuli/Poger\_DPPC/tests\_gmx\_versions/)[Samuli, you may want to add this Figure+Legend+Comment 
%o a stable link (Figshare?), let me know]. 
%Of note, some tests (Miguel Machuqueiro, personnal communication) with the same RF scheme i) on simulations with proteins 
%(thus with explicit hydrogens) showed that this setup is still unstable and eventually leads to crashes with versions 4.6.6 and 4.6.7 (and so does version 4.5.*).
Thus we decided to use only the PME setup ii) for computing the order parameter since it gives stable results regardless of the GROMACS version. We obtained an area per 
lipid of 0.615~nm$^2$, below 0.648 nm$^2$ found by the original authors with their PME setup (see~\cite{poger12}). We explained that by the fact that we used 
a 1.4~nm for the LJ cut-off whereas a value of 1.0~nm was used in the original publication.

@@

\subsubsection{CHARMM36-UA}
A hydrated bilayer consisting of 128 DLPC lipids and 3840 water molecules is modeled by the force field of Lee and co-workers~\cite{lee14}.
This force field is a combination of the all-atom CHARMM36 force-field~\cite{klauda10} and the united-atom Berger model~\cite{berger97}. 
The non-bonded interactions are calculated using an atom-based switching function with short and long cut-offs of 0.8 and 1.2~nm~\cite{lee14}. 
Long range electrostatic interactions are implemented using the particle-particle particle-mesh solver with a relative accuracy of $10^{-4}$. The system 
is first equilibrated for 30~ns in the NP$\gamma$T ensemble (Nos\'{e}--Hoover~\cite{nose84,hoover85} style thermostat and barostat with anisotropic pressure coupling) 
at 323~K and 1~bar with timestep of 1~fs. The following 20~ns of dynamics are taken for calculation of configurational averages. 
\todo{The simulation times given currently contradict with the Table~\ref{systems}.}
Simulations were 
carried out by using the LAMMPS package~\cite{plimpton95}. %( http://doi.org/10.1006/jcph.1995.1039 )., the input files are available ( http://dx.doi.org/10.5281/zenodo.13821 ).

@@

\subsection{Dihedral angle distributions as a function of cholesterol in CHARMM36}
\begin{figure*}[]
  \centering
%  \includegraphics[width=17.2cm]{dihsCHOLcharmm.pdf}
  \caption{\label{dihsCHOLcharmm}
    The effect of cholesterol content on the glycerol backbone and choline dihedral angles in CHARMM36 model.}
\end{figure*}

>>>>

  \begin{figure}[]
  \centering
  \includegraphics[width=8.6cm]{POPCstructure.eps}

@@

\sout {Berger-POPC-07~\cite{ollila07a}}\todoi{Not included in the Fig. 2, to be removed?}          &   POPC & 128 & 7290  & 323  & 56 & 56  & ? & SI \\

@@

CHARMM36\cite{klauda10}      & POPC   & 128 &  5120    & 303 & 150 & 100  & ? \todoi{Permanent link in progress by Hubert Santuz}   & SI \\
CHARMM36\cite{klauda10}       & DPPC   & 72  &  2189 & 323 & 30 & 25  & ?\todoi{Samuli, put to Zenodo}  & SI \\
CHARMM36\cite{klauda10}\todoi{Not included in the Fig. 2. Should we remove or include?}        & DOPC   & 128 &  5120    & 303 & 150 & 100  & ?\todoi{Permanent link in progress by Hubert Santuz}  & SI \\

@@

Poger\cite{poger10}             & DPPC  & 128 & 5841 & 323 & 2$\times$100 & 2$\times$50 & \cite{pogerFILES}\todoi{Permanent link in progress by Patrick Fuchs} & SI \\

@@

\subsection{Full hydration: Experimental order parameters for the glycerol backbone and headgroup}\label{experiments}
The specific deuteration of $\alpha$-, $\beta$- and g$_3$- segments of dipalmitoylphosphatidylcholine (DPPC) has been successful, 
allowing the order parameter measurements for these segments by $^2$H--NMR~\cite{gally75,brown77,brown78,akutsu81}.
In addition, the order parameters for all glycerol backbone and choline headgroup segments in egg yolk lecithin~\cite{hong95a},
1,2-dimyristoyl-sn-glycero-3-phosphocholine (DMPC)~\cite{hong95b,gross97,dvinskikh05a}, 
1,2-dioleoyl-sn-glycero-3-phosphocholine (DOPC)~\cite{warschawski05} and POPC~\cite{warschawski05,ferreira13}
have been measured with several different implementations of $^{13}$C NMR experiments.
The experimental absolute values of glycerol backbone and choline order parameters from various publications are shown in Fig.~\ref{HGorderparameters}.
\begin{figure}[]
  \centering
  \includegraphics[width=8.6cm]{comparison3.pdf}
%\todo{DONE}\\
%The data in this figure should be updated to be consistent with the systems listed in Table~\ref{systems}. At least, Tjornhammar et al and CHARMM-UA values are missing. Some results in the figure are outdated due to longer simulations reported etc.} \\
\todo{Markus Miettinen suggested that we should consider making one more figure where only experimental data would be shown and that would be discussed in Section~\ref{experiments}} \\
\todo{There has been some discussion in the blog that maybe we should include the signs and stereospecific labeling here as well, and that maybe this figure should be clarified. My view is that for the main conclusion, i.e. CHARMM36, GAFFlipid and MacRog are the best here, this format is good enough. However, if someone wants to improve this and/or collect all the data with the signs and/or stereospecific labeling it would be even better. There is also very interesting discussion about interactive figure.}
  \caption{\label{HGorderparameters}
  Order parameteres from simulations listed in Table~\ref{systems} and experiments for glycerol and choline groups.
  The experimental values were taken from the following publications: DMPC 303~K from \cite{gross97}, DMPC 314~K from \cite{dvinskikh05a}, DPPC 310K from \cite{gally75}, 
  DPPC 323~K from \cite{akutsu81}, and POPC 298~K \cite{ferreira13}.
  The vertical bars shown for some of the computational values are not error bars, but demonstrate that for 
  these systems we had two data sets; the ends of the bars mark the values from the two sets, and the dot marks their measurement-time-weighted average. 
  The H{\"o}gberg et al. force field in Table~\ref{systems} is abbreviated in this figure as HoNiLy, Tj{\"o}rnhammar et al. as TjoEdh, CHARMM36-UA as C36UA,
  Rabinovic et al. as RabLyu and Ulmschneider and Ulmschneider as UlmUlm.
} 
\end{figure}

@@

By combining the experimental information of the sign~\cite{hong95a,hong95b,gross97} and the stereospecific
measurements~\cite{gally81} with the accurate absolute value measurements from various techniques~\cite{gally75,akutsu81,gross97,dvinskikh05a,ferreira13},
the most detailed experimentally available order parameter information for the glycerol backbone and choline segments of POPC is obtained.
These data are shown in Fig.~\ref{HGorderparameters2}.
\begin{figure*}[]
  \centering
  \includegraphics[width=8.6cm]{HGorderparameters5.eps} \\
  \todo{Markus Miettinen suggested that we should consider ``spreading'' the values similarly to the figure~\ref{HGorderparameters}} \\
  \caption{\label{HGorderparameters2}
  Order parameteres from simulations with Berger-POPC-07, CHARMM36, GAFFlipid and MacRog force fields together with experimental values for POPC glycerol and choline groups.
  The magnitudes for experimental order parameters are taken from Ferreira et al.~\cite{ferreira13}, the signs are based on the measurements by Hong et al.~\cite{hong95a,hong95b} 
  and Gross et al.~\cite{gross97}, and the R/S labeling is based in the measurements by Gally et al.~\cite{gally81}.
} 
\end{figure*}

@@

The dihedral angle distributions for the  g$_3$-g$_2$-g$_1$-O(\textit{sn}-1) dihedral calculated from different models are
shown in Fig.~\ref{dihDISTS}. The distribution is qualitatively different for the Berger-POPC-07 model, showing a maximum in 
the gauche$^+$-conformation (60$^o$) compared to all the other models showing a maximum in the anti-conformation (180$^o$).
The distributions in all the other models have the same general features, the main difference being that the
fraction of configurations in the gauche$^-$-conformation (-60$^o$) is zero for the MacRog, detectable for the CHARMM36 and
equally large to the gauche$^+$ fraction in GAFFlipid. From the results we conclude that most likely the wrongly sampled
dihedral angle for the g$_2$-g$_1$ bond explains the significant discrepancy to the experimental order parameters
for the g$_1$ segment in the Berger-POPC-07 model (Fig.~\ref{HGorderparameters2}). 
In conclusion, models preferring the anti conformation for this dihedral give more realistic order parameters and
this in agreement with previous crystal structure and $^1$H NMR studies~\cite{hauser80,hauser81,hauser81b,hauser88,pascher92,marsh06}.
\begin{figure}[]
  \centering
  \includegraphics[width=8.6cm]{g1-g2_Cdihs2.eps}
  \caption{\label{dihDISTS}
    Dihedral angle distributions for g$_3$-g$_2$-g$_1$-O(\textit{sn}-1) dihedral from different models (POPC bilayer in full hydration).
      } 
\end{figure}

@@

%Interestingly, the probability of the gaughe conformations correlates with the order parameter difference between the $\beta$ and $\alpha$ segments:
%the larger the gaughe fraction the larger the order parameter difference. This suggestion together with the results in 
%Fig.~\ref{HGorderparameters2} would indicate that the correct gaughe-trans fraction for N-$\beta$-$\alpha$-O dihedral is 
%larger than in GAFFlipid but smaller than in CHARMM36. This happens to be quite close to the gaughe-trans fraction in
%the Berger model. However, there is significant forking and numerical values are off from the experiments in the Berger
%model suggesting that it has some other inaccuracies.
\begin{figure}[]
  \centering
  \includegraphics[width=8.6cm]{a-bDIHS2.eps}
  \caption{\label{dihDISTS2}
    Dihedral angle distributions for N-$\beta$-$\alpha$-O dihedral from different models (POPC bilayer in full hydration).
  } 
\end{figure}

@@

\subsubsection{Phospholipid bilayer with low hydration level}
The experimental order parameters available in the literature~\cite{dvinskikh05b,ulrich94,bechinger91} 
for the glycerol backbone and choline as a function of hydration level are shown in Fig.~\ref{ordPhydr}. 
The independently reported values for choline segments are in good agreement with each other (despite 
slight differences in temperature and acyl chain composition),
showing increase for both segments with decreasing hydration level. It should be noted that only 
absolute values were measured in the original experiments~\cite{dvinskikh05b,ulrich94,bechinger91}, but
we have included the signs measured in separate studies~\cite{hong95a,hong95b,gross97}. 
Consequently, the $\beta$ order parameter with negative sign actually increases with dehydration 
since the absolute value decreases~\cite{dvinskikh05b,ulrich94,bechinger91}.
Slight decrease for the glycerol backbone g$_3$- and g$_2$- order parameters were observed with dehydration, 
while g$_1$ remained practically unchanged~\cite{dvinskikh05b}.
\begin{figure}[]
  \centering
  \includegraphics[width=8.6cm]{OrderParameterDEHYDpresentation.eps}
\todo{Several people think that this plot should be clarified.}
  \caption{\label{ordPhydr}
    The effect of dehydration on glycerol and choline order parameters in experiments.
    The magnitudes of order parameters are measured for DMPC ($^{13}$C NMR) at 314~K~\cite{dvinskikh05b}, 
    for POPC ($^2$H NMR) at 296~K~\cite{bechinger91} and for DOPC ($^2$H NMR) at 303~K~\cite{ulrich94}. 
    The signs are based on the measurements by Hong et al.~\cite{hong95a,hong95b} 
    and Gross et al.~\cite{gross97}.
  }
\end{figure}

@@

The qualitative agreement with experiments in all simulation models for the $\alpha$ and $\beta$ order parameters  
as a function of hydration indicates that the structural response of the choline headgroup to dehydration is somewhat realistic
despite the unrealistic structures at full hydration. 
The most likely explanation is that the choline group
orients more parallel to the membrane plane with dehydration due to restricted interlamellar space. 
Indeed, the P--N (phoshate phosphorus to choline nitrogen) vector angle with membrane normal shows an increase for
all models as a function of dehydration in Fig.~\ref{PNangle}.
However, the qualitative agreement in the lipid response to dehydration does not guarantee the correct 
free energy landscape if the simulation model has an incorrect structure. The influence of this issue to 
dehydration energetics studied with simulations~\cite{eun09,schneck12} is left for future studies.
\begin{figure}[]
  \centering
  \includegraphics[width=8.6cm]{PNangles.eps}

@@

The choline headgroup and glycerol backbone order parameters for POPC measured by $^{13}$C NMR~\cite{ferreira13} and DPPC choline order parameters 
measured by $^{2}$H NMR~\cite{brown78} are shown in Fig.~\ref{ordPchol} as a function of cholesterol content.
The agreement between different experimental results is again very good, showing only very modest changes in 
the choline order parameters as a function of cholesterol content. It should be noted, however, that very small
changes are measurable with high resolution $^{2}$H NMR experiments
and cholesterol causes a measurable increase in the $\beta$ order parameter and a forking in the $\alpha$ order
parameter~\cite{brown78}. These effects are, however, so small that they are barely visible in the scale used in Fig.~\ref{ordPchol}.
Further, the effects of cholesterol on the glycerol backbone order parameters for POPC from $^{13}$C NMR experiment~\cite{ferreira13} 
are in good agreement with the results for the phosphatidylethanolamine (PE) measured by $^{2}$H NMR~\cite{ghosh82}.
These results further support the idea that the glycerol backbone structural behaviour is independent of the
headgroup composition~\cite{gally81} and that the headgroup stucture is independent of the acyl chain region content unless
charges are present~\cite{scherer87}.
\begin{figure}[]
  \centering
  \includegraphics[width=8.6cm]{OrderParameterCHOLexpVSsim.eps}
  \caption{\label{ordPchol}
    The effect of cholesterol content on the glycerol backbone and choline order parameters in experiments~\cite{brown78,ferreira13} and simulations
    with the Berger-POPC-07/H\"oltje-CHOL-13, CHARMM36 and MacRog force fields. The signs in the experimental values are based on the measurements by Hong et al.~\cite{hong95a,hong95b} 
    and Gross et al.~\cite{gross97}.  Most order parameters from Berger-POPC-07/H\"oltje-CHOL-13 model for g$_1$ are beoynd the y-axis scale.}
\end{figure}

@@

Since Poger lipids come from the GROMOS force field, it is important to note that GROMOS uses the RF scheme for computing electrostatics (this is the method used for the 
force field parameterization). Using setup i) based on RF, we were able to reproduce the results (i.e. area per lipid value of 0.63~nm$^2$) from the original work only with 
GROMACS versions 4.0.X and earlier (the original authors~\cite{poger10} used GROMACS version 3.3.3). When switching to versions 4.5.X and above, the area per lipid dropped to below 0.58~nm$^2$. 
The GROMACS developers were contacted and a redmine issue opened (\texttt{http://redmine.gromacs.org/issues/1400}). The difference comes from the new Trotter decomposition 
introduced in versions 4.5.X. A fix has been introduced in version 4.6.6 that allows a recovery of an area per lipid value of 0.615~nm$^2$. The results in terms of area per lipid using the different 
GROMACS versions are available at~\cite{fuchs14}\todoi{Permanent link in progress by Patrick Fuchs}.
% http://www.dsimb.inserm.fr/~fuchs/project\_Samuli/Poger\_DPPC/tests\_gmx\_versions/)[Samuli, you may want to add this Figure+Legend+Comment 
%o a stable link (Figshare?), let me know]. 
%Of note, some tests (Miguel Machuqueiro, personnal communication) with the same RF scheme i) on simulations with proteins 
%(thus with explicit hydrogens) showed that this setup is still unstable and eventually leads to crashes with versions 4.6.6 and 4.6.7 (and so does version 4.5.*).
Thus we decided to use only the PME setup ii) for computing the order parameter since it gives stable results regardless of the GROMACS version. We obtained an area per 
lipid of 0.615~nm$^2$, below 0.648 nm$^2$ found by the original authors with their PME setup (see~\cite{poger12}). We explained that by the fact that we used 
a 1.4~nm for the LJ cut-off whereas a value of 1.0~nm was used in the original publication.

@@

\subsubsection{CHARMM36-UA}
A hydrated bilayer consisting of 128 DLPC lipids and 3840 water molecules is modeled by the force field of Lee and co-workers~\cite{lee14}.
This force field is a combination of the all-atom CHARMM36 force-field~\cite{klauda10} and the united-atom Berger model~\cite{berger97}. 
The non-bonded interactions are calculated using an atom-based switching function with short and long cut-offs of 0.8 and 1.2~nm~\cite{lee14}. 
Long range electrostatic interactions are implemented using the particle-particle particle-mesh solver with a relative accuracy of $10^{-4}$. The system 
is first equilibrated for 30~ns in the NP$\gamma$T ensemble (Nos\'{e}--Hoover~\cite{nose84,hoover85} style thermostat and barostat with anisotropic pressure coupling) 
at 323~K and 1~bar with timestep of 1~fs. The following 20~ns of dynamics are taken for calculation of configurational averages. 
Simulations were carried out by using the LAMMPS package~\cite{plimpton95}. %( http://doi.org/10.1006/jcph.1995.1039 )., the input files are available ( http://dx.doi.org/10.5281/zenodo.13821 ).

@@

\subsection{Dihedral angle distributions as a function of cholesterol in CHARMM36}
\begin{figure*}[]
  \centering
  \includegraphics[width=17.2cm]{dihsCHOLcharmm.pdf}
  \caption{\label{dihsCHOLcharmm}
    The effect of cholesterol content on the glycerol backbone and choline dihedral angles in CHARMM36 model.}
\end{figure*}

\subsection{General issues to think of about the manuscript}
\todo{Should we emphasize more the excellent collection of simulation trajectories collected to Zenodo?}
\todo{Maybe we should mention that CHARMM is significanlty more computationally expensive than other all-atom models.
This seems evident for a person who has ran CHARMM and other all-atom models with the same computer. However, there
may be some people who have never used any other model than CHARMM. In this case it may be difficult to realize.
Would anyone have a good bencmark about this? I have assumed that the reason for the slowness are the extra Lennart-Jones
interactionsi in water, however, I have never looked into it. Would someone have some hard facts on this?}
